Appendix 1: sdmTMB statistical model description

### 2021-12-21 doc/appendix-model.Rmd
Add to model appendix

@@

@@

AR(1) spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. 
They are defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. 
The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.

@@

@@

$$
\operatorname{Normal} \left( \mu, \phi \right)
$$
where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of standard deviation not variance).
The variance is $\phi^2$.

@@

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

@@

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$ [@hilbe2011].

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

@@

Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is fit as an unconstrained variable.

@@

where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
The parameter $\nu$ controls the smoothness of the covariance function.
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

Internally, the random fields are assumed to follow a Gaussian Markov random field. For example, the spatial random field case:

$$
\boldsymbol{\omega} \sim \mathrm{MVN}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}\right)
$$
where $\boldsymbol{Q}$ is a sparse precision matrix.

@@

$$
\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega}
$$
The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
Three non-zero elements on each row define the weight of the neighbouring 3 knot locations.
The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$.

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on and off with the boolean `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance REF, and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$

@@

In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes). 
SPDE models allow for barriers to be included via barrier meshes. 
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers. 
See `?sdmTMB::add_barrier_mesh` for examples.

TODO DETAILS ON  METHOD HERE

@@

@@

Much of the guidance around diagnostics and `glmmTMB` also applies to `sdmTMB`, e.g. [https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html). Warnings of non-positive-definite Hessian matrices (accompanied by parameters with NAs for standard errors) often mean models are improperly specified given the data used. Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`) -- more iterations and smaller tolerance thresholds ensure that the optimal solution is found, but more evaluations translates into longer computation time. Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlmbinb_loops` as an argument to `sdmTMB()`, where each call is initialized at the best parameter vector from the previous iteration. Alternatively, `sdmTMB()` may be called repeatedly, using different random seeds.

>>>>

Internally, the random fields follow a Gaussian Markov random field (GMRF):

$$
\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
$$
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.

@@

Similarly to the spatial random fields, these spatiotemporal random fields (and all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):

$$
\boldsymbol{\epsilon}_t \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
$$

@@

First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. 
They are defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. 
The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ term scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. 
The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.

@@

Here we describe the main observation families that are available in sdmTMB and comment on their parametrization, statistical properties, utility, and code representation in sdmTMB.

@@

$$
\operatorname{Normal} \left( \mu, \phi^2 \right)
$$
where $\mu$ is the mean and $\phi$ is the standard deviation.
The variance is $\phi^2$.

@@

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

@@

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$ [@hilbe2011].
Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

@@

Internally, $p$ is transformed from $\mathrm{logit}^{-1} (\texttt{thetaf}) + 1$ to constrain it between 1 and 2 and \texttt{thetaf} is estimated as an unconstrained variable.

Example: `family = tweedie(link = "log")`

@@

where $\tau^2$ controls the spatial variance,
$\nu$ controls the smoothness,
$\Gamma$ represents the Gamma function,
$d_{jk}$ represents the distance between locations $s_j$ and $s_k$,
$K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ controls the spatial correlation range.
The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

@@

$$
\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega},
$$
where $\boldsymbol{\omega}^*$ represents the values of the spatial random fields at the observed locations or predicted data locations.
The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $s$.
The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$:

$$
\boldsymbol{\epsilon}_t^* = \boldsymbol{A} \boldsymbol{\epsilon}_t.
$$

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance REF, and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\textbf{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$

@@

In some cases the spatial domain of interest may be complex and bordered by land or water (e.g., coastlines, islands, lakes). 
SPDE models allow for barriers to be included via barrier meshes [@bakka2019]. 
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`. 
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers. 
See `?sdmTMB::add_barrier_mesh` for examples.

The barrier mesh implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019]. 
For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assume that the range was 2 km across the barrier. 
This would let the spatial correlation decay 5 times faster with distance. 
From experimentation, values around 0.1 or 0.2 seem to work well but values much lower than 0.1 can result in convergence issues.

The website <https://haakonbakkagit.github.io/btopic128.html> provides an illustration with INLA. The implementation within TMB was borrowed from code written by Olav Nikolai Breivik and Hans Skaug at <https://github.com/skaug/tmb-case-studies>.

@@

# Optimization

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`)---a greater number of iterations and smaller tolerance thresholds increase the chance that the optimal solution is found, but more evaluations translates into longer computation time.
Additionally, multiple function calls to `stats::nlminb()` may aid convergence; we include `nlminb_loops` as an argument to the control list in `sdmTMB()` (e.g., `sdmTMB(..., control = sdmTMBcontrol(nlmbinb_loops = 2))`), where each call is initialized at the best parameter vector from the previous iteration.
Similarly, additional optimization with a Newton solver (`stats::optimHess()`) can be implemented by adding `newton_loops` in the control list.
If a model is already fit, the function `sdmTMB::run_extra_optimization()` can run additional optimization loops with either routine to further reduce the maximum gradient.

@@

Much of the guidance around diagnostics and glmmTMB also applies to `sdmTMB`, e.g. [https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html). 
Warnings of non-positive-definite Hessian matrices (accompanied by parameters with `NA`s for standard errors) often mean models are improperly specified given the data.
Standard errors can be observed in the output of `print.sdmTMB()` or by checking `fit$sd_report`.
The maximum gradient of the marginal likelihood with respect to fixed-effect parameters can be checked by inspecting (`fit$gradients`).
Guidance varies, but the maximum gradient should likely be at least $< 0.001$ before assuming the fitting routine is consistent with convergence.
If maximum gradients are already relatively small, they can often be reduced further with additional optimization calls beginning at the previous best parameter vector as described above with `sdmTMB::run_extra_optimization()`.
